Semi-Supervised Factored Logistic Regression for High-Dimensional Neuroimaging Data

### 2015-06-05 paper_nips/paper.tex
removed non-centered from PCA-autoencoder description
grammar

Both these neurobiological properties can be conjointly exploited in an
approach that is
\textit{mixed} (i.e., using rest and task data),
\textit{factored} (i.e., performing network decomposition), and
\textit{multi-task} (i.e., capitalize on neural representations shared across
mental operations).
%
The integration of brain-network discovery into
supervised classification can yield a semi-supervised learning
framework. The most relevant
neurobiological structure should hence be identified
for the prediction problem at hand.
%
Autoencoders suggest themselves because they can emulate
variants of most unsupervised learning algorithms,
including PCA, SPCA, and ICA \cite{hinton06}.
\begin{wrapfigure}{r}{0.40\textwidth}
  \centering
  %\captionsetup{labelformat=nonumber}
    \includegraphics[width=0.40\textwidth]{figures/figure1.png}
  \vspace{-0.7cm}
  \caption {\textbf{Model architecture.}
  Linear autoencoders find an optimized compression of 79,941 brain voxels
  into $n$ unknown activity patterns by improving reconstruction from them.
  The decomposition matrix equates with the bottleneck of
  a factored logistic regression.
  Supervised multi-class learning on task data ($X_{task}$)
  can thus be guided by
  unsupervised decomposition of rest data ($X_{rest}$).
  }
\end{wrapfigure}
%
Autoencoders (AE)
are layered learning models
that condense the input data to
local and global 
representations
via reconstruction under compression prior.
%
% XXX Michael: In fact, they behave like a non-centered PCA, 
% XXX which amounts to just doing an SVD
% XXX which btw is NOT a convex optimization problem, although easy to 
% XXX calculate and with a global optimum.
% XXX Autoencoders are also not one-layered by default. Let's make sure that
% XXX we don't trip up by getting the technicalities wrong and getting 
% XXX reviewers annoyed at us because of it.
% DanilO: cool thanks. That is the stuff, we need.
They behave like a (truncated, non-centered) PCA
in case of one linear hidden layer and a squared error loss
\cite{baldi1989neural}.
% This architecture yields an optimization objective
% with unique global minimum.
Autoencoders behave like a SPCA if shrinkage terms are added to the
model weights in the optimization objective.
Moreover, they have the characteristics of an ICA in case of tied weights
and adding a nonlinear convex function
at the first layer \cite{le2011ica}.
% XXX What convex transformation? Why is that relevant? Convexity is only
% XXX relevant for the loss
% Danilo: it's about adding a neural-network-like nonconvex function for
% first-layer activation
These authors further demonstrated that ICA, sparse autoencoders, and 
sparse coding are mathematically equivalent
under mild conditions.
%
Thus, autoencoders may flexibly project the neuroimaging data
onto the main directions of variation.

In the present investigation,
a linear autoencoder will be fit to
(unlabeled) rest data and
integrated as a rank-reducing bottleneck
into a multinomial
logistic regression fit to
(labeled) task data.
We can then solve the unsupervised data representation and the
supervised classification, previously studied in isolation,
in an identical statistical learning problem.
%
From the perspective of dictionary learning, the first layer represents
% XXX Michael: This is the crucial difference: The first layer is an analysis
% XXX operation: it extracts information in some way. It may be extracting
% XXX activations for basis functions, in which case it would be the pseudo-
% XXX inverse of those basis functions. But unless they are orthogonal, these
% XXX weights definitely don't represent basis functions themselves. It is 
% XXX the second layer of the autoencoder that represents basis functions. If
% XXX weights are tied, they are one and the same thing, with interesting
% XXX implications. But in general, only the second layer represents a basis.
projectors to the discovered set of basis functions
which are linearly combined 
by the second layer to perform predictions \cite{olshausen96}.
% XXX Michael: ^ This needs to be changed around: This first layer obtains
% XXX activations, the second layer weights are basis functions which are
% XXX multiplied to these activations
%
Neurobiologically, this allows 
delineating a low-dimensional manifold
of brain network patterns and then 
distinguishing mental tasks
by their most discriminative linear combinations.
%
Theoretically, a reduction in model variance should be achieved by
resting-state autoencoders that
privilege the most neurobiologically
valid models in the hypothesis set.
%
Practically, neuroimaging research is frequently suffering from
data scarcity. This limits the set of representations that can be
extracted from GLM analyses based on few participants.
%
We therefore contribute a computational framework that
1) analyzes many problems simultaneously
(thus finds shared representations by ``multi-task learning'')
and
2) exploits unlabeled data,
as they span a space of meaningful configurations.

>>>>

Both these neurobiological properties can be conjointly exploited in an
approach that is
\textit{mixed} (i.e., using rest and task data),
\textit{factored} (i.e., performing network decomposition), and
\textit{multi-task} (i.e., capitalize on neural representations shared across
mental operations).
%
The integration of brain-network discovery into
supervised classification can yield a semi-supervised learning
framework. The most relevant
neurobiological structure should hence be identified
for the prediction problem at hand.
%
Autoencoders suggest themselves because they can emulate
variants of most unsupervised learning algorithms,
including PCA, SPCA, and ICA \cite{hinton06}.
\begin{wrapfigure}{r}{0.40\textwidth}
  \centering
  %\captionsetup{labelformat=nonumber}
    \includegraphics[width=0.40\textwidth]{figures/figure1.png}
  \vspace{-0.7cm}
  \caption {\textbf{Model architecture.}
  Linear autoencoders find an optimized compression of 79,941 brain voxels
  into $n$ unknown activity patterns by improving reconstruction from them.
  The decomposition matrix equates with the bottleneck of
  a factored logistic regression.
  Supervised multi-class learning on task data ($X_{task}$)
  can thus be guided by
  unsupervised decomposition of rest data ($X_{rest}$).
  }
\end{wrapfigure}
%
Autoencoders (AE)
are layered learning models
that condense the input data to
local and global 
representations
via reconstruction under compression prior.
%
% XXX Michael: In fact, they behave like a non-centered PCA, 
% XXX which amounts to just doing an SVD
% XXX which btw is NOT a convex optimization problem, although easy to 
% XXX calculate and with a global optimum.
% XXX Autoencoders are also not one-layered by default. Let's make sure that
% XXX we don't trip up by getting the technicalities wrong and getting 
% XXX reviewers annoyed at us because of it.
% DanilO: cool thanks. That is the stuff, we need.
They behave like a (truncated) PCA
in case of one linear hidden layer and a squared error loss
\cite{baldi1989neural}.
% This architecture yields an optimization objective
% with unique global minimum.
Autoencoders behave like a SPCA if shrinkage terms are added to the
model weights in the optimization objective.
Moreover, they have the characteristics of an ICA in case of tied weights
and adding a nonlinear convex function
at the first layer \cite{le2011ica}.
% XXX What convex transformation? Why is that relevant? Convexity is only
% XXX relevant for the loss
% Danilo: it's about adding a neural-network-like nonconvex function for
% first-layer activation
These authors further demonstrated that ICA, sparse autoencoders, and 
sparse coding are mathematically equivalent
under mild conditions.
%
Thus, autoencoders may flexibly project the neuroimaging data
onto the main directions of variation.

In the present investigation,
a linear autoencoder will be fit to
(unlabeled) rest data and
integrated as a rank-reducing bottleneck
into a multinomial
logistic regression fit to
(labeled) task data.
We can then solve the unsupervised data representation and the
supervised classification, previously studied in isolation,
in an identical statistical learning problem.
%
From the perspective of dictionary learning, the first layer represents
% XXX Michael: This is the crucial difference: The first layer is an analysis
% XXX operation: it extracts information in some way. It may be extracting
% XXX activations for basis functions, in which case it would be the pseudo-
% XXX inverse of those basis functions. But unless they are orthogonal, these
% XXX weights definitely don't represent basis functions themselves. It is 
% XXX the second layer of the autoencoder that represents basis functions. If
% XXX weights are tied, they are one and the same thing, with interesting
% XXX implications. But in general, only the second layer represents a basis.
projectors to the discovered set of basis functions
which are linearly combined 
by the second layer to perform predictions \cite{olshausen96}.
% XXX Michael: ^ This needs to be changed around: This first layer obtains
% XXX activations, the second layer weights are basis functions which are
% XXX multiplied to these activations
%
Neurobiologically, this allows 
delineating a low-dimensional manifold
of brain network patterns and then 
distinguishing mental tasks
by their most discriminative linear combinations.
%
Theoretically, a reduction in model variance should be achieved by
resting-state autoencoders that
privilege the most neurobiologically
valid models in the hypothesis set.
%
Practically, neuroimaging research frequently suffers from
data scarcity. This limits the set of representations that can be
extracted from GLM analyses based on few participants.
%
We therefore contribute a computational framework that
1) analyzes many problems simultaneously
(thus finds shared representations by ``multi-task learning'')
and
2) exploits unlabeled data,
as they span a space of meaningful configurations.
